---
title: "PA1_template.Rmd"
output: html_document
---

## 1. Loading and preprocessing the data

1. Load the data
```{r results = "hide"}
activity <- read.csv("./activity.csv")
```

2. Process/transform the data into a format suitable for the analysis
- convert date column from factor to date class
- remove all rows with NA's
```{r}
activity$date <- as.Date(activity$date)
activitySub <- na.omit(activity)
```

## 2. What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
```{r message=FALSE}
library(dplyr)
```

```{r}
stepsTotal <- summarize(group_by(activitySub, date), sum = sum(steps))
stepsTotal
```

2. Make a histogram of the total number of steps taken each day.
```{r}
hist(stepsTotal$sum)
```

Here is an alternative method to creating the histogram with ggplot2:
```{r}
library(ggplot2)
ggplot(stepsTotal, aes(x=sum)) + geom_histogram(binwidth = 1000)
```

3. Calculate and report the mean and median of the total number of steps taken per day.
```{r}
mean <- mean(stepsTotal$sum)
median <- median(stepsTotal$sum)
mean; median
```

## 3. What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days
```{r}
intervalAverage <- summarize(group_by(activitySub, interval), averageSteps = mean(steps))
intervalAverage
with(intervalAverage, plot(interval, averageSteps, type = "l"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
intervalAverage[which.max(intervalAverage$average),]
```

## 4. Imputing missing values
1. Calculate and report the total number of missing values in the dataset.  
```{r}
sum(is.na(activity$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset: *I will replace NA's with the mean value for that 5 minute interval.*

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r message=FALSE}
library(plyr)
```

```{r}
impute.mean <- function(x){
        replace(x, is.na(x), mean(x, na.rm = TRUE))
        }
activityImputed <- ddply(activity, ~ interval, transform, steps = impute.mean(steps))
detach("package:plyr", unload=TRUE)
```

These are the first rows of the data set before imputing NA's:
```{r}
head(activity)
```

These are the first rows of the data set after imputing NA's.  You can see that the number of steps in the first row has changed from an NA in the original data set to 1.71, which is the interval 0 average, as observed in step 3 above.
```{r}
head(activityImputed)
```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.  

Based on the graphs and mean/median calculated below, it does not appear that replacing the NA's with imputed values had significant impact.

```{r}
stepsTotal2 <- summarize(group_by(activityImputed, date), sum = sum(steps))
hist(stepsTotal2$sum)
library(ggplot2)
ggplot(stepsTotal2, aes(x=sum)) + geom_histogram(binwidth = 1000)
mean <- mean(stepsTotal2$sum)
median <- median(stepsTotal2$sum)
mean; median
```

## 5. Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels, indicating whether a given date is a weekday or weekend day.
```{r}
activityImputed$weekday <- weekdays(activityImputed$date)
activityImputed$dayType <- ifelse(activityImputed$weekday %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
activityImputed$dayType <- as.factor(activityImputed$dayType)
head(activityImputed)
class(activityImputed$dayType)
```

2. Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.

```{r}
intervalAverage3 <- summarize(group_by(activityImputed, interval, dayType), aveSteps = mean(steps))
intervalAverage3
library(lattice)
xyplot(aveSteps ~ interval | dayType, data = intervalAverage3, type = "l")
```

